---
output: github_document
format: gfm
---
# arcgisutils <img src="man/figures/logo.svg" align="right" height="139" alt="" />

<!-- badges: start -->
[![R-CMD-check](https://github.com/R-ArcGIS/arcgisutils/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/R-ArcGIS/arcgisutils/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/arcgisutils)](https://CRAN.R-project.org/package=arcgisutils)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


arcgisutils is designed as the backbone of the [`{arcgis}`](https://github.com/r-arcgis/arcgis) meta-package.

arcgisutils is a developer oriented package that provides the basic functions to build R packages that work with ArcGIS Location Services. It provides functionality for:

- authorization, 
- Esri JSON construction and parsing, 
- structuring ArcGIS REST API requests,
- Esri type conversions,
- among other utilities

## Installation

Install arcgisutils from CRAN.

``` r
install.packages("arcgisutils", repos = "https://r-arcgis.r-universe.dev")
```

Or, you can install the development version of arcgisutils [r-universe](https://r-arcgis.r-universe.dev/) with:

``` r
install.packages("arcgisutils", repos = "https://r-arcgis.r-universe.dev")
```

### Authorization 

Authorization tokens are provided through the functions `auth_code()`, `auth_client()`, `auth_user()`, and `auth_binding()`. Additional token validation functions are provided via `refresh_token()` and `validate_or_refresh_token()`. 

Tokens are managed in a session based cache using `set_arc_token()` and `unset_arc_token()`. They are fetched using `arc_token()`. Here is a minimal example:


```{r}
library(arcgisutils)

tkn <- auth_client()

set_arc_token(tkn)

arc_token()
```

Alternatively, tokens can be set based on a key-value pair. 

```{r}
set_arc_token("A" = tkn, "B" = tkn)
```

And fetched based on their name via

```{r}
arc_token("A")
```

### Standardized Requests 

The function `arc_base_req()` is used to create a standardized `httr2` request object. It handles authorization tokens and sets a user agent.

```{r}
host <- arc_host() # use arcgis.com by default

arc_base_req(host)
```

### Esri JSON

There are also a number of utility functions for creating and parsing Esri JSON. For example we can create a list that represent an Esri `FeatureSet` using `as_featurset()` directly from an `sf` object. To convert to json, it is recommended to use `jsonify::to_json(x, unbox = TRUE)`.

```{r}
library(sf)

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
nc_json <- as_featureset(nc)

str(nc_json, 1)
```

Alternatively, you can use the set of functions with the `_esri_` infix to directly create the json. See the [Esri geometry reference page](https://r.esri.com/arcgisutils/reference/esri_geometry.html) for more on how the conversion functions work. 

Additionally, sf's `crs` object can be converted to a [`spatialReference`](https://developers.arcgis.com/documentation/common-data-types/geometry-objects.htm#GUID-DFF0E738-5A42-40BC-A811-ACCB5814BABC) JSON object using `validate_crs()`.

```{r}
validate_crs(27700)
```





