---
output: github_document
format: gfm
---
# arcgisutils <img src="man/figures/logo.svg" align="right" height="139" alt="" />

<!-- badges: start -->
[![R-CMD-check](https://github.com/R-ArcGIS/arcgisutils/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/R-ArcGIS/arcgisutils/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/arcgisutils)](https://CRAN.R-project.org/package=arcgisutils)
[![extendr](https://img.shields.io/badge/extendr-^0.8.0-276DC2)](https://extendr.rs/extendr/extendr_api/)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(pillar)
```


arcgisutils is the foundational infrastructure package that powers the R-ArcGIS Bridge Data and Location Service ecosystem. It provides sophisticated, production-ready tools for interacting with ArcGIS Online, ArcGIS Enterprise, and ArcGIS Platform via their REST APIs.

## Key Capabilities

**🔐 Comprehensive Authentication**:

- Multiple OAuth2 workflows (`auth_code()`, `auth_client()`) 
- API key and legacy token support
- Automatic token refresh and validation
- Integration with ArcGIS Pro via `arcgisbinding`

**🌐 Portal Integration**: 

- Advanced content search with filtering, sorting, and pagination
- User, group, and organization metadata management
- Portal item discovery and content management workflows

**⚙️ Geoprocessing Services**: 

- Support for the geoprocessing service framework built upon R6 and S7
- Enables users to call their own custom geoprocessing services or build on top of existing services
- R6-based job management (`arc_gp_job`) with real-time status tracking and built-in result parsing

**📄 Esri JSON Ecosystem**: 

- Bidirectional conversion between R spatial data and Esri JSON formats
- Support for FeatureSets, geometry objects, field definitions, and spatial reference systems
- Optimized parsing with automatic `sf` integration

**🛠️ Developer Utilities**: 

- Standardized HTTP client (`arc_base_req()`, `arc_paginate_req()`)
- Robust error detection and user-friendly error messages
- URL parsing, service introspection, and metadata extraction

## Installation

`{arcgisutils}` is part of the `{arcgis}` metapackage, which provides the complete R-ArcGIS Bridge toolkit. For most users, installing the metapackage is recommended:

```r
install.packages("arcgis")
```

You can also install `{arcgisutils}` individually from CRAN:

```r
install.packages("arcgisutils")
```

To install the development version:

```r
pak::pak("r-arcgis/arcgisutils")
```

### Authentication

Authorization tokens are provided through the functions `auth_code()`, `auth_client()`, `auth_user()`, `auth_key()`, and `auth_binding()`. Additional token validation functions are provided via `refresh_token()` and `validate_or_refresh_token()`. 

`auth_code()` can be used for integrating into Shiny applications, for example, to have individual users log in. We recommend using `auth_key()` for authenticating in non-interactive environments (for example scheduled scripts or deployments).

Tokens are managed using `set_arc_token()` and `unset_arc_token()`. They are fetched using `arc_token()`. `set_arc_token()` can set the token globally or set multiple named environments. Here is a minimal example:

```{r, eval=TRUE}
library(arcgisutils)
```

```{r eval=FALSE}
key <- auth_key()
set_arc_token(key)
```

Alternatively, tokens can be set based on a key-value pair for multiple environments:

```{r eval=FALSE}
set_arc_token("production" = prod_token, "development" = dev_token)
```

And fetched based on their name via

```{r eval=FALSE}
arc_token("production")
```

### Portal Integration

Search and discover content across your ArcGIS organization:

```{r}
# Search for feature services containing "crime" data
crime_items <- search_items(
  query = "crime", 
  item_type = "Feature Service",
  max_pages = 1
)

crime_items
```

```{r}
# Get detailed item information for a portal item
arc_item(crime_items$id[1])
```

### Developer Utilities

Always use `arc_base_req()` as this will handle setting the user agent and authorization token. The function creates a standardized `httr2` request object:

```{r}
# defaults to arcgis.com
host <- arc_host() 

req <- arc_base_req(host)
req
```

To handle paginated services and requests use `arc_paginate_req()` to automatically handle fetching pages.


### Esri JSON

There are also a number of utility functions for creating and parsing Esri JSON. For example we can create an Esri `FeatureSet` json string using `as_esri_featureset()` directly from an `sf` object. 

```{r, message=FALSE, warning = FALSE}
library(sf)

# load the NC SIDS dataset and extract centroids
# of the first few rows
nc <- system.file("shape/nc.shp", package = "sf") |> 
  st_read(quiet = TRUE) |> 
  st_centroid() 

# convert to json
nc_json <- as_esri_featureset(nc[1:2, 1:3])

jsonify::pretty_json(nc_json)
```


Feature set json can also be parsed using `parse_esri_json()`.

```{r}
parse_esri_json(nc_json)
```


Additionally, sf's `crs` object can be converted to a [`spatialReference`](https://developers.arcgis.com/documentation/common-data-types/geometry-objects.htm#GUID-DFF0E738-5A42-40BC-A811-ACCB5814BABC) JSON object using `validate_crs()`. Convert these to json with `yyjsonr` or `jsonify`.

```{r}
crs <- validate_crs(27700)
jsonify::pretty_json(crs, unbox = TRUE)
```

### Geoprocessing Services

The geoprocessing service framework is completely supported in `{arcgisutils}`. Here we combine the functionality of the geoprocessing job framework with utilities such as `as_esri_featureset()` to call the [Trace DownStream Elevation Service](https://developers.arcgis.com/rest/elevation-analysis/trace-downstream/)

```{r}
trace_downstream <- function(
  input_points,
  point_id_field = NULL,
  resolution = NULL,
  generalize = FALSE,
  token = arc_token()
) {

  # create a list of parameters
  params <- compact(list(
    InputPoints = as_esri_featureset(input_points),
    PointIdField = point_id_field,
    DataSourceResolution = resolution,
    Generalize = as.character(generalize),
    f = "json"
  ))

  service_url <- "https://hydro.arcgis.com/arcgis/rest/services/Tools/Hydrology/GPServer/TraceDownstream"
  arc_gp_job$new(
    base_url = service_url,
    params = params,
    result_fn = parse_gp_feature_record_set,
    token
  )
}
```


This new function can be called to start a new job: 

```{r}
# create input points
input_points <- st_sfc(
  st_point(c(-159.548936, 21.955888)),
  crs = 4326
)

# initialze an empty job
job <- trace_downstream(
  input_points,
  token = auth_user()
)

# start the job
job$start()
```

Jobs run asynchronously so we can check the status with `job$status`

```{r include = FALSE}
Sys.sleep(0.5)
```

```{r}
job$status
```

Then, when the job is complete, we can fetch the results applying the result function which is `parse_gp_feature_record_set()` in this case. 


```{r include=FALSE}
Sys.sleep(5)
```

```{r}
job$results

# store and view the results
res <- job$results
res

# plot the resultant geometry
plot(st_geometry(res$geometry))
```

## Learn More

To learn more about the R-ArcGIS Bridge project visit the [developer documentation](https://developers.arcgis.com/r-bridge/).